---
title: "Script for DWS"
format: pdf
editor: visual
---

## Preparation of the Data

Libraries

```{r}
library(dplyr)
library(tidyr)
library(lcmm)
library(ggplot2)
```

### ADNI data TADPOLE challenge

```{r}
tad <- read.csv(file="~/Seafile/_NOT_SHARED_RESTE/ANALYSES_ADNI/TADPOLE_D1_D2.csv")

names(tad)[grepl("ST29", names(tad))]

## create variables of volumes
## HIPP: 
tad$hippFSX <- tad$ST29SV_UCSFFSX_11_02_15_UCSFFSX51_08_01_16 / 1000
tad$hippFSL <- tad$ST29SV_UCSFFSL_02_01_16_UCSFFSL51ALL_08_01_16 /1000

## Histograms of the volumes
par(mfrow=c(1,2))
hist(tad$hippFSX)
hist(tad$hippFSL)
```

### Creation of the dataset

Time in Study and recentered age at baseline

Selection of the observations with both pipelines

```{r}
tadS <- tad %>%   mutate(
    EXAMDATE    = as.Date(EXAMDATE, format = "%Y-%m-%d"),
    EXAMDATE_bl = as.Date(EXAMDATE_bl, format = "%Y-%m-%d"),
    TiS=as.numeric(EXAMDATE - EXAMDATE_bl)/365.25,
    AGE75 = AGE - 75) %>% filter(!is.na(hippFSL)&!is.na(hippFSX))
```

# description of the sample

Number of repeated measures of MRI

```{r}
tadS <- tadS %>% group_by(RID) %>% mutate(
    nmes   = n()) %>%ungroup()

Desc <- tadS %>% group_by(RID) %>% filter(TiS==0)

quantile(Desc$nmes)

```

### Plots by patterns

#### Number of MRI

```{r}
dataGG <- tadS %>%
  select(RID, nmes, hippFSL, hippFSX) %>%
  pivot_longer(cols = c(hippFSL, hippFSX), names_to = "hipp", values_to = "levels")

ggplot(dataGG, aes(x = factor(nmes), y = levels, fill = hipp)) +
  geom_violin(position = position_dodge(width = 0.8), alpha = 0.7) +
  labs(
    x = "Number of measures (nmes)",
    y = "levels",
    fill = "hipp",
    title = "Distribution of hippL et hippX by nmes pattern"
  ) +
  theme_minimal()

```

#### Sex

```{r}
dataGG <- tadS %>%
  select(RID, PTGENDER, hippFSL, hippFSX) %>%
  pivot_longer(cols = c(hippFSL, hippFSX), names_to = "hipp", values_to = "levels")

ggplot(dataGG, aes(x = factor(PTGENDER), y = levels, fill = hipp)) +
  geom_violin(position = position_dodge(width = 0.8), alpha = 0.7) +
  labs(
    x = "Sex",
    y = "levels",
    fill = "hipp",
    title = "Distribution of hippL et hippX"
  ) +
  theme_minimal()
```

#### AGE75

```{r}
tadS <- tadS %>%
  mutate(
    AGE75cat = cut(
      AGE75,
      breaks = c(seq(-20, 10, by = 5), 16),
      include.lowest = TRUE,
      right = TRUE
    )
  )
summary(tadS$AGE75cat)
```

```{r}
dataGG <- tadS %>%
  select(RID, AGE75cat, hippFSL, hippFSX) %>%
  pivot_longer(cols = c(hippFSL, hippFSX), names_to = "hipp", values_to = "levels")

ggplot(dataGG, aes(x = factor(AGE75cat), y = levels, fill = hipp)) +
  geom_violin(position = position_dodge(width = 0.8), alpha = 0.7) +
  labs(
    x = "Sex",
    y = "levels",
    fill = "hipp",
    title = "Distribution of hippL et hippX"
  ) +
  theme_minimal()
```

# Linear mixed models

Time in Study as timescale + adjustment for Age0

```{r}
library(lcmm)
mhippL <- hlme(hippFSL ~ TiS + AGE75, random=~TiS,subject="RID",data=tadS)
mhippX <- hlme(hippFSX ~ TiS+ AGE75, random=~TiS,subject="RID",data=tadS)

summary(mhippL)
summary(mhippX)
```

```{r}
REhipp <- data.frame(RID=mhippX$predRE$RID, 
                     intX=mhippX$predRE$intercept,
                     intL= mhippL$predRE$intercept,
                     TiSX=mhippX$predRE$TiS,
                     TiSL=mhippL$predRE$TiS)


length(mhippL$predRE$intercept)

pint <- ggplot(REhipp)+
  geom_point(aes(x=intX,y=intL))+
  geom_abline(slope = 1,intercept = 0,color="blue")

pint

pTiS <- ggplot(REhipp)+
  geom_point(aes(x=TiSX,y=TiSL))+
  geom_abline(slope = 1,intercept = 0,color="blue")

pTiS
```

```{r}
m <- lm(TiSL ~ TiSX, data=REhipp)
summary(m)
```

#### with 2 measures max for longi? 

Vincent has some issues with the error variance at 0

```{r}
tadS2 <- tadS %>% filter(nmes <=2)
  
mhippL2 <- hlme(hippFSL ~ TiS + AGE75, random=~TiS,subject="RID",data=tadS2)
mhippX2 <- hlme(hippFSX ~ TiS+ AGE75, random=~TiS,subject="RID",data=tadS2)

summary(mhippL2)
summary(mhippX2)
```

```{r}
REhipp2 <- data.frame(RID=mhippX2$predRE$RID, 
                     intX=mhippX2$predRE$intercept,
                     intL= mhippL2$predRE$intercept,
                     TiSX=mhippX2$predRE$TiS,
                     TiSL=mhippL2$predRE$TiS)


length(mhippL2$predRE$intercept)

pint <- ggplot(REhipp2)+
  geom_point(aes(x=intX,y=intL))+
  geom_abline(slope = 1,intercept = 0,color="blue")

pint

pTiS <- ggplot(REhipp2)+
  geom_point(aes(x=TiSX,y=TiSL))+
  geom_abline(slope = 1,intercept = 0,color="blue")

pTiS
```

```{r}
m2 <- lm(TiSL ~ TiSX, data=REhipp2)
summary(m2)
```
