---
title: "Script for DWS"
editor: visual
format:
  html:
    toc: true
    toc-depth: 3          # profondeur des titres inclus (par défaut 3)
    toc-location: left    # left, right, ou body
    number-sections: true # numérote les sections (souvent utilisé avec toc)
---

# Preparation of the Data

Libraries

```{r}
library(dplyr)
library(tidyr)
library(lcmm)
library(ggplot2)
library(simex)
library(nlme)
library(survival)
library(patchwork)   # used to arrange multiple ggplot objects into one panel
```

## ADNI data TADPOLE challenge

```{r}
tad <- read.csv(file="~/Seafile/_NOT_SHARED_RESTE/ANALYSES_ADNI/TADPOLE_D1_D2.csv")
Dict <- read.csv("~/Seafile/_NOT_SHARED_RESTE/ANALYSES_ADNI/TADPOLE_D1_D2_Dict.csv")

names(tad)[grepl("ST29", names(tad))]

names(tad)[grepl("ST88", names(tad))]

### Renamme columns for cortical thicknesses ----
names_col<-c("RID", "AGE","PTGENDER","PTEDUCAT","APOE4","EXAMDATE","EXAMDATE_bl",
             "DX_bl","DX",
             "MMSE","RAVLT_immediate","CDRSB", "CDRSB_bl",
             "ICV", "Hippocampus",#"TOTAL_WMH",
             "AV45",
             names(tad)[grep("TA_UCSFFS", names(tad))],#cortical thickness
             "ST29SV_UCSFFSX_11_02_15_UCSFFSX51_08_01_16",
             "ST29SV_UCSFFSL_02_01_16_UCSFFSL51ALL_08_01_16",
             "ST88SV_UCSFFSL_02_01_16_UCSFFSL51ALL_08_01_16",
             "ST88SV_UCSFFSX_11_02_15_UCSFFSX51_08_01_16"   ,
             "ST128SV_UCSFFSX_11_02_15_UCSFFSX51_08_01_16",
             "ST128SV_UCSFFSL_02_01_16_UCSFFSL51ALL_08_01_16")

# Renames columns of cortical thicknesses : Region_Cross or Region_Long
names_CT <- names(tad)[grep("TA_UCSFFS", names(tad))]

idx <- match(names_CT, Dict$FLDNAME)

names_new <- ifelse(
  grepl("Cross-Sectional FreeSurfer", Dict$CRFNAME[idx]),
  paste0(
    sub("^Cortical Thickness Average of ", "", Dict$TEXT[idx]),
    "_Cross"
  ),
  paste0(
    sub("^Cortical Thickness Average of ", "", Dict$TEXT[idx]),
    "_Long"
  )
)

names_new <- c(names_new,"hippLeftFSX","hippLeftFSL","hippRightFSX","hippRightFSL", "WMHFSX","WMHFSL")
tad <- tad[, names_col]
names(tad)[17:dim(tad)[2]] <- names_new


## create variables of volumes
## HIPP Left: 
tad$hippLeftFSX <- tad$hippLeftFSX / 1000
tad$hippLeftFSL <- tad$hippLeftFSL /1000
## HIPP Right: 
tad$hippRightFSX <- tad$hippRightFSX / 1000
tad$hippRightFSL <- tad$hippRightFSL /1000


## Histograms of the volumes
par(mfrow=c(1,2))
hist(tad$hippLeftFSX)
hist(tad$hippLeftFSL)

```

## Creation of the dataset

Time in Study and recentered age at baseline

Selection of the observations with both pipelines

```{r}
tadS <- tad %>%   mutate(
    EXAMDATE    = as.Date(EXAMDATE, format = "%Y-%m-%d"),
    EXAMDATE_bl = as.Date(EXAMDATE_bl, format = "%Y-%m-%d"),
    TiS=as.numeric(EXAMDATE - EXAMDATE_bl)/365.25,
    AGE75 = AGE - 75) %>% filter(!is.na(hippLeftFSL)&!is.na(hippLeftFSX))
```

# description of the sample

Number of repeated measures of MRI

```{r}
tadS <- tadS %>% group_by(RID) %>% mutate(
    nmes   = n()) %>%ungroup()

Desc <- tadS %>% group_by(RID) %>% filter(TiS==0)

quantile(Desc$nmes)

```

## Plots by patterns

### Number of MRI

```{r}
dataGG <- tadS %>%
  select(RID, nmes, hippLeftFSL, hippLeftFSX) %>%
  pivot_longer(cols = c(hippLeftFSL, hippLeftFSX), names_to = "hipp", values_to = "levels")

ggplot(dataGG, aes(x = factor(nmes), y = levels, fill = hipp)) +
  geom_violin(position = position_dodge(width = 0.8), alpha = 0.7) +
  labs(
    x = "Number of measures (nmes)",
    y = "Hipp volume",
    fill = "pipeline",
    title = "Distribution of hippL et hippX by nmes pattern"
  ) +
  theme_minimal()
```

### Number of MRI on data at baseline


```{r}
dataGG <- tadS %>% filter(TiS==0) %>%
  select(RID, nmes, hippLeftFSL, hippLeftFSX) %>%
  pivot_longer(cols = c(hippLeftFSL, hippLeftFSX), names_to = "hipp", values_to = "levels")

ggplot(dataGG, aes(x = factor(nmes), y = levels, fill = hipp)) +
  geom_violin(position = position_dodge(width = 0.8), alpha = 0.7) +
  labs(
    x = "Number of measures (nmes)",
    y = "Hipp volume",
    fill = "pipeline",
    title = "Distribution of hippL et hippX by nmes pattern"
  ) +
  theme_minimal()
```


### Sex

```{r}
dataGG <- tadS %>%
  select(RID, PTGENDER, hippLeftFSL, hippLeftFSX) %>%
  pivot_longer(cols = c(hippLeftFSL, hippLeftFSX), names_to = "hipp", values_to = "levels")

ggplot(dataGG, aes(x = factor(PTGENDER), y = levels, fill = hipp)) +
  geom_violin(position = position_dodge(width = 0.8), alpha = 0.7) +
  labs(
    x = "Sex",
    y = "Hipp volume",
    fill = "pipeline",
    title = "Distribution of hippL et hippX"
  ) +
  theme_minimal()
```

### AGE75

```{r}
tadS <- tadS %>%
  mutate(
    AGE75cat = cut(
      AGE75,
      breaks = c(seq(-20, 10, by = 5), 16),
      include.lowest = TRUE,
      right = TRUE
    )
  )
summary(tadS$AGE75cat)
```

```{r}
dataGG <- tadS %>%
  select(RID, AGE75cat, hippLeftFSL, hippLeftFSX) %>%
  pivot_longer(cols = c(hippLeftFSL, hippLeftFSX), names_to = "hipp", values_to = "levels")

ggplot(dataGG, aes(x = factor(AGE75cat), y = levels, fill = hipp)) +
  geom_violin(position = position_dodge(width = 0.8), alpha = 0.7) +
  labs(
    x = "age",
    y = "levels",
    fill = "hipp",
    title = "Distribution of hippL et hippX"
  ) +
  theme_minimal()
```



## Bland-Altman plot 


The function first

```{r}
# -----------------------------------------------------------------------------
# Compute Bland-Altman statistics (mean, difference, bias, limits of agreement)
# -----------------------------------------------------------------------------
bland_altman_stats <- function(x, y) {
  # x, y: numeric vectors of equal length (method 1 and method 2 measurements)
  mean_xy   <- (x + y) / 2
  diff_xy   <- x - y
  bias      <- mean(diff_xy, na.rm = TRUE)
  sd_diff   <- sd(diff_xy, na.rm = TRUE)
  loa_lower <- bias - 1.96 * sd_diff
  loa_upper <- bias + 1.96 * sd_diff

  data.frame(mean = mean_xy, diff = diff_xy,
             bias = bias, loa_lower = loa_lower, loa_upper = loa_upper)
}

# -----------------------------------------------------------------------------
# Build a single Bland-Altman ggplot for one variable
# -----------------------------------------------------------------------------
bland_altman_plot <- function(data, var_name, suffix1 = "_1", suffix2 = "_2") {
  # data     : data.frame containing the paired columns
  # var_name : base variable name (without suffix)
  # suffix1  : suffix identifying method 1 column
  # suffix2  : suffix identifying method 2 column

  col1 <- paste0(var_name, suffix1)
  col2 <- paste0(var_name, suffix2)

  if (!all(c(col1, col2) %in% names(data))) {
    stop(paste("Columns", col1, "and/or", col2, "not found in data"))
  }

  ba_data <- bland_altman_stats(data[[col1]], data[[col2]])

  ggplot(ba_data, aes(x = mean, y = diff)) +
    geom_point(alpha = 0.6, color = "steelblue") +
    geom_hline(aes(yintercept = bias), color = "black", linewidth = 0.6) +
    geom_hline(aes(yintercept = loa_lower), color = "red", linetype = "dashed") +
    geom_hline(aes(yintercept = loa_upper), color = "red", linetype = "dashed") +
    labs(title = var_name,
         x = "Mean of X and L",
         y = "Difference (X - L)") +
    theme_minimal(base_size = 10)
}

# -----------------------------------------------------------------------------
# Build a panel of Bland-Altman plots for a vector of variable names
# -----------------------------------------------------------------------------
bland_altman_panel <- function(data, vars, suffix1 = "_1", suffix2 = "_2", ncol = 2) {
  # data    : data.frame containing all the paired columns
  # vars    : character vector of base variable names (the input vector)
  # suffix1 : suffix identifying method 1 columns (default "_1")
  # suffix2 : suffix identifying method 2 columns (default "_2")
  # ncol    : number of columns in the panel layout

  plots <- lapply(vars, function(v) bland_altman_plot(data, v, suffix1, suffix2))
  wrap_plots(plots, ncol = ncol)
}
```

Its application


```{r}
tadS$logWMHFSX <- log(tadS$WMHFSX)
tadS$logWMHFSL <- log(tadS$WMHFSL)
listRegions <- c("hippLeft","hippRight", "WMH","logWMH")


p <- bland_altman_panel(tadS, listRegions,
                         suffix1 = "FSX", suffix2 = "FSL", ncol = 2)

print(p)
```

# MRI as outcome : lnear mixed models

Time in Study as timescale + adjustment for Age0

```{r}
library(lcmm)
mhippL <- hlme(hippLeftFSL ~ TiS + AGE75, random=~TiS,subject="RID",data=tadS)
mhippX <- hlme(hippLeftFSX ~ TiS+ AGE75, random=~TiS,subject="RID",data=tadS)

summary(mhippL)
summary(mhippX)
```

```{r}
REhipp <- data.frame(RID=mhippX$predRE$RID, 
                     intX=mhippX$predRE$intercept,
                     intL= mhippL$predRE$intercept,
                     TiSX=mhippX$predRE$TiS,
                     TiSL=mhippL$predRE$TiS)


length(mhippL$predRE$intercept)

pint <- ggplot(REhipp)+
  geom_point(aes(x=intX,y=intL))+
  geom_abline(slope = 1,intercept = 0,color="blue")

pint

pTiS <- ggplot(REhipp)+
  geom_point(aes(x=TiSX,y=TiSL))+
  geom_abline(slope = 1,intercept = 0,color="blue")

pTiS
```

```{r}
m <- lm(TiSL ~ TiSX, data=REhipp)
summary(m)
```

## with 2 measures max for longi? 

Vincent has some issues with the error variance at 0

```{r}
tadS2 <- tadS %>% filter(nmes <=2)
  
mhippL2 <- hlme(hippLeftFSL ~ TiS + AGE75, random=~TiS,subject="RID",data=tadS2)
mhippX2 <- hlme(hippLeftFSX ~ TiS+ AGE75, random=~TiS,subject="RID",data=tadS2)

summary(mhippL2)
summary(mhippX2)
```

```{r}
REhipp2 <- data.frame(RID=mhippX2$predRE$RID, 
                     intX=mhippX2$predRE$intercept,
                     intL= mhippL2$predRE$intercept,
                     TiSX=mhippX2$predRE$TiS,
                     TiSL=mhippL2$predRE$TiS)


length(mhippL2$predRE$intercept)

pint <- ggplot(REhipp2)+
  geom_point(aes(x=intX,y=intL))+
  geom_abline(slope = 1,intercept = 0,color="blue")

pint

pTiS <- ggplot(REhipp2)+
  geom_point(aes(x=TiSX,y=TiSL))+
  geom_abline(slope = 1,intercept = 0,color="blue")

pTiS
```

```{r}
m2 <- lm(TiSL ~ TiSX, data=REhipp2)
summary(m2)
```


## Heteroskedasticity


```{r}
library(nlme)
tadS <- tadS %>% mutate(nmes2=ifelse(nmes==1,1,ifelse(nmes==2,2,3)))

mhippLnh <- lme(hippLeftFSL ~ TiS + AGE75, random=~TiS | RID,data=tadS)
mhippLh <- lme(hippLeftFSL ~ TiS + AGE75, random=~TiS | RID,data=tadS,
  weights = varIdent(form = ~ 1 | nmes2)
)

mhippXnh <- lme(hippLeftFSX ~ TiS + AGE75, random=~TiS | RID,data=tadS)
mhippXh <- lme(hippLeftFSX ~ TiS + AGE75, random=~TiS | RID,data=tadS,
  weights = varIdent(form = ~ 1 | nmes2)
)

anova(mhippLnh,mhippLh)
anova(mhippXnh,mhippXh)

summary(mhippL)
summary(mhippX)
```



# MRI as Covariate: SIMEX for hipp as the covariate? 

#### 0. The data 

```{r}
tadS$EP_VolL <- tadS$hippLeftFSL
tadS$EP_VolX <- tadS$hippLeftFSX
tadS$Cog <- tadS$RAVLT_forgetting
```

#### The Pipeline

1. I create a model to retrieve an estimate of the residual error of my volume
2. I create the naive primary model of interest 
3. I use SIMEX


#### 1. Model for the residual error

```{r}
mhippL <- hlme(hippLeftFSL ~ TiS + AGE75, random=~TiS,subject="RID",data=tadS)
mhippX <- hlme(hippLeftFSX ~ TiS+ AGE75, random=~TiS,subject="RID",data=tadS)

sdEpsHippL <- abs(mhippL$best[length(mhippL$best)])
sdEpsHippX <- abs(mhippX$best[length(mhippX$best)])
```
#### I test with a cross-sectional outcome

I use cognition at baseline to test the package with a lm model

For longitudinal pipeline

```{r}
# data baseline
tadS0 <- tadS %>% filter(TiS==0)

# naive model
LMM0 <- lm( Cog ~ EP_VolL, data = tadS,na.action = na.omit)
summary(LMM0)

# simex
res_simex <- simex(LMM0,
                    SIMEXvariable     = "EP_VolL",
                    measurement.error = sdEpsHippL,
                    lambda            = c(0.5, 1, 1.5, 2,3,4),
                    B                 = 200,
                    fitting.method    = "quadratic",   
                    asymptotic        = FALSE)         # FALSE => variance par jackknife
summary(res_simex)
plot(res_simex)
```

For a cross-sectional pipeline

```{r}
# data baseline
tadS0 <- tadS %>% filter(TiS==0)

# naive model
LMM0 <- lm( Cog ~ EP_VolX, data = tadS,na.action = na.omit)
summary(LMM0)

# simex
res_simex <- simex(LMM0,
                    SIMEXvariable     = "EP_VolX",
                    measurement.error = sdEpsHippX,
                    lambda            = c(0.5, 1, 1.5, 2,3,4),
                    B                 = 200,
                    fitting.method    = "quadratic",   
                    asymptotic        = FALSE)         # FALSE => variance par jackknife
summary(res_simex)
plot(res_simex)
```


#### I test now with a Time-to-event outcome 

```{r}
tadSurv <- tadS %>%
  group_by(RID) %>%
  summarise(
    time_dementia = if (any(grepl("Dementia", DX, ignore.case = TRUE), na.rm = TRUE))
                        min(TiS[grepl("Dementia", DX, ignore.case = TRUE)], na.rm = TRUE)
                     else NA_real_,
    time_last_visit = max(TiS, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    TimeEvent   = if_else(!is.na(time_dementia), time_dementia, time_last_visit),
    StatusEvent = if_else(!is.na(time_dementia), 1L, 0L)
  )

dim(tadSurv)

tadSurv <- merge(tadSurv,tadS0,by="RID")

# naive primary model
Cox0 <- coxph(Surv(TimeEvent,StatusEvent) ~ EP_VolL + AGE75, data = tadSurv,
              model=TRUE)
summary(Cox0)

# simex
set.seed(123)
res_simex <- simex(Cox0,
                    SIMEXvariable     = "EP_VolL",
                    measurement.error = sdEpsHippL,
                    lambda            = c(0.5, 1, 1.5, 2,3,4,5,10),
                    B                 = 200,
                    fitting.method    = "quadratic",   
                    asymptotic        = FALSE)         # FALSE => variance par jackknife
summary(res_simex)
plot(res_simex)
```


